---
title: "Age-wage profile"
author: "Jesús Yancy - Juan Otalora"
date: "2025-09-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set working directory
username <- Sys.getenv("USER")

if (username == "jesusyancy"){
  path <- "/Users/jesusyancy/Documents/MECA/BDML/TALLERES/TALLER 1"
} else if(username == ''){
  path <- ""
}
setwd(path)
```

```{r}
library(pacman)
```

```{r}
# Instalar paquetes
p_load(rio, # import/export data
       tidyverse, # tidy-data
       skimr, # summary data
       visdat, # visualizing missing data
       corrplot, # Correlation Plots 
       stargazer) # tables/output to TEX.
```

```{r}
df <- read_csv("db_clean.csv")
```

## 3. Age-wage profile. 

A great deal of evidence in labor economics suggests that the typical worker’s age-wage profile has a predictable path: “Wages tend to be low when the worker is young; they rise as the worker ages, peaking at about age 50; and the wage rate tends to remain stable or decline slightly after age 50”.

In this subsection we are going to estimate the Age-wage profile profile for the individuals in this sample:

$$ log(w) = β_1 + β_2Age+ β_3Age^2 + u $$

```{r}
df <- df %>%
  mutate(age2 = age^2)
```

```{r}
model_age <- lm(Log_Total_hour_salary ~ age + age2, data = df)

stargazer(model_age, type = "text")
```

Peak age

```{r}
# Extraer coeficientes
coefs <- coef(model_age)
b2 <- coefs["age"]
b3 <- coefs["age2"]

peak_age <- -b2 / (2 * b3)
peak_age

peak_salary <- coefs["(Intercept)"] + b2*peak_age + b3*peak_age^2
exp(peak_salary)

```


```{r}
library(ggplot2)

# Coeficientes del modelo original
coefs <- coef(model_age)
b1 <- coefs["(Intercept)"]
b2 <- coefs["age"]
b3 <- coefs["age2"]

# Peak age estimado
peak_age <- -b2 / (2 * b3)
peak_salary <- b1 + b2*peak_age + b3*peak_age^2

# Crear un rango de edades para graficar la curva
age_seq <- seq(min(df$age, na.rm = TRUE), max(df$age, na.rm = TRUE), length.out = 200)
pred_salary <- b1 + b2*age_seq + b3*age_seq^2

# Dataframe para graficar
plot_data <- data.frame(age = age_seq, salary = pred_salary)

# Gráfico
ggplot(plot_data, aes(x = age, y = salary)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(aes(x = peak_age, y = peak_salary), color = "red", size = 3) +
  geom_vline(xintercept = peak_age, linetype = "dashed", color = "red") +
  labs(
    title = "Curva cuadrática de salario horario vs. edad",
    subtitle = paste0("Peak age ≈ ", round(peak_age, 1), " años"),
    x = "Edad",
    y = "Salario horario (log o nivel)"
  ) +
  theme_minimal()

```

## Calcular los intervalos de confianza del peak age con bootstrap

```{r}
# Se define la función que recibe los datos y un vector de índices que boot genera en cada subset de la muestra con reemplazo.
# La función devuelve el peak age para cada subset.
peak_fn <- function(data, index) {
  fit <- lm(Log_Total_hour_salary ~ age + age2, data = data, subset = index) # Estima el modelo con cada subset
  coefs <- coef(fit)
  -coefs["age"] / (2 * coefs["age2"]) # Calcular el peak age de cada subset
}

# Se prueba en la muestra original
peak_fn(df, 1:nrow(df))

# bootstrap con 1000 réplicas
set.seed(123)
boot_out <- boot(df, statistic = peak_fn, R = 1000)

# resultados
boot_out$t0          # peak age de la muestra original
boot_out             # resumen del bootstrap

# intervalos de confianza
boot.ci(boot_out, type = "perc")
```














Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
