---
title: "gender-earnings-gap"
author: "Juan José Rojas & Francisco Soler"
date: "2025-09-04"
output: html_document
---

```{r}
install.packages("pacman")
library(pacman)

p_load(rio, # Import/export data.
       tidyverse, # Tidy-data.
       stargazer, # Descriptive statistics.
       gt, # Descriptive statistics.
       gtsummary,
       caret, # For predictive model assessment.
       gridExtra, # Arrange plots.
       skimr # Summarize data.
       )
```

# Database load

```{r}
# Working directory setting
username <- Sys.getenv("USERNAME")
# Dynamic filepath
ruta_datos <- file.path("C:/Users", username, "Documents",
                        "MEcA - Uniandes/2025-2/Big Data y Machine Learning/PS1_Group4/data")
# Load original database                  
db<- read_csv(file.path(ruta_datos, "db_clean1.csv"))
dim(db) # database dimensions
```

# Unconditional Gender-Wage GAP

$$ log(w) = β_1 + β_2Female + u $$

```{r}
# Create indicator of female
db <- db |>
  mutate(bin_female = 1 - sex)
db$bin_female <- factor(db$bin_female)

# Run the model 
model_sex <- lm(Log_Total_hour_salary ~ bin_female, data = db)
# View the coefficients
stargazer(model_sex, type = "text")

```

The unconditional regression indicates that women earn, on average, about 4.7 percent less per hour than men. This gap is estimated without controlling for any observable characteristics, so it reflects the raw difference in average hourly wages between the two groups. In other words, the coefficient captures the unconditional gender wage gap in log-wage terms.

# Equal Pay for Equal Work?

We need to control for: - Similar worker characteristics: Education (edu years, max level, college), age or experience.

-   Similar job characteristics: formality (formal), size of the business (sizeFirm, microEmpresa), occupation (oficio, relab)

$$ log(w) = \beta_1 + \beta_2Female + \beta_3Education + \beta_4Age + \beta_5Age^2 + \beta_5Formality + \beta_6Business Size + \gamma_1Oficio + u $$

$$ log(w) = \beta_1 + \beta_2Female + \beta_3Education + \beta_4Experience + \beta_5Age^2 + \beta_5 + u $$

```{r}
db <- db %>%
  mutate(age2 = age^2)

db$oficio <- factor(db$oficio)
db$formal <- factor(db$formal)
db$microEmpresa <- factor(db$microEmpresa)
db$relab <- factor(db$relab)
  
# Individual characteristics (edu_years + age + age2)
model_equality1 <- lm(Log_Total_hour_salary ~ bin_female + edu_years + age + age2, data = db)

# Individual + Firm nature
model_equality2 <- lm(Log_Total_hour_salary ~ bin_female + edu_years + age + age2 + formal +  sizeFirm + relab, data = db)

# Individual + Firm nature + occupation
model_equality3 <- lm(Log_Total_hour_salary ~ bin_female + edu_years + age + age2 + formal +  sizeFirm + relab + oficio, data = db)


# View the coefficients
stargazer(model_sex, model_equality1, model_equality2, model_equality3, type = "text", out = "tabla_modelo.txt")

```

```{r}
db %>%
  mutate(sex = factor(sex, levels = c(1,0), labels = c("Men","Women"))) %>%
  select(sex, edu_years, Log_Total_hour_salary, age) %>%
  tbl_summary(by = sex,
              statistic = all_continuous() ~ "{mean} ({sd})",
              digits = all_continuous() ~ 2) %>%
  add_n() %>%           # agrega N por grupo
  add_p()               # prueba de diferencia (opcional)

```